---
title: "First Lines of Code"
author: "Michael Barrows"
date: "`r Sys.Date()`"
output:
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: slides.css
---

## Adgenda 

1. Basic Calculations
2. Assignment
3. Subsetting
4. Functions

# 1. Basic Calculations

## Just Like a Calculator

```{r echo = T}
1 + 1 # Basic Operations
200 * 1 / (9 + 1)^2 - 1 # Order of Operations
```

- Works just like a calculator
- Follows PEMDAS 
- New line = New calculation

## Your Turn

- Find the product of 134 and 25
- Find the sum of 43 and 27. Then divide it by 7 in one line
- Find the square root of 60 (hint: sqrt(__))

## Solutions 
```{r}
134 * 25
(43 + 27) / 7
sqrt(60)
```

# 2. Assignment

## Save values for later

<div class = "left">
```{r echo = T, eval = T}
sqrt_2 <- sqrt(2)
sqrt_2 * 5

sqrt_2by5 <- sqrt_2 * 5
sqrt_2by5 + 1

sqrt_2 <- sqrt(3)
```

</div>

<div class = "right">
  - use the left arrow 
  - variable on left, value on right
  - reuse values without recalculating
  - use variables to create new variable
      * must be declared before they can be used
  - values stored in MEMORY
  - can be overwritten
</div>

## Your Turn

- assign 6 to "x"
- assign x + 5 to "y"
- rewrite x to be 10
- assign y + x to "z"

## Solutions
```{r}
x <- 6
y <- x + 5
x <- 10
z <- y + x
```


# 3. Subsetting

## Array

<div class = "left">
```{r echo = T, eval = F}
new_array <- c(1,2,3,4) #Create
new_array[1] #First Element
new_array[1:3] #Elements 1-3
new_array[-2] #Not Element 2
```

</div>

<div class = "right">
  - Create with "c()"
  - Subset with "\[\]"
  - 1 indexed
  - use "-" as *not* operator
</div>

## Your Turn

- Create an array with numbers 11-20
- Select the value "13"
- Select the 5th element
- [Challenge] Select all but the 8th and 10th element

## Solutions
```{r eval = F}
test_array <- c(11:20)
test_array[3]
# OR
test_array[test_array == 13]
test_array[5]
test_array[-c(8,10)]
```


## Matrix

- Two dimensional array
- All entries are the same data type
- Subset with "matrix[row,column]"

##
<div class = "left">
```{r echo = T}
new_matrix <- matrix(1:9,3,3)
new_matrix
new_matrix[1,2]
new_matrix[1] 
```

</div>

<div class = "right">
```{r}
new_matrix[1:3] 
new_matrix[1:2,3] 
new_matrix[,1:2]
```

</div>


## Your Turn

```{r eval = T, echo=F}
matrix(1:9,3,3)
```

Use the same matrix as above to:

- find the value in the 2nd row, 3rd column
- find the values in the 3rd column
- find the 5th value 

## Solutions

```{r}
mat <- matrix(1:9,3,3)
mat[2,3]

mat[,3]

mat[5]
```

## Data Frame

<div class = "left">
```{r echo = T, eval = F}
head(mtcars, 6)
mtcars[1] #not first value
mtcars[2,4]
mtcars[2,]
mtcars["cyl"]
```

</div>

<div class = "right">
  - Similar to matrix 
  - Can call columns by name
  - Single value in brackets is now columns
  - Can mix data types
</div>

## Your Turn

Use the mtcars database to: 

- Subset the 6th column
- find the 5th value in the 2nd row.

## Solutions

```{r}
mtcars[6]
# OR
mtcars[,6]

mtcars[2,5]
```

## List

<div class = "left">
```{r echo = T, eval = F}
list1 <- list(mtcars, 1:20, 
              letters[1:13])
str(list1)
list1[1]
list1[[1]][1,2]

results <- lm(mpg~wt, data=mtcars)
str(results)
```

</div>

<div class = "right">

- Groups of objects
- All above data types can be placed in a list
- Think XML or JSON data
- subsetting syntax: "name[[outer]][inner]"
- Alternatively: "name$outer[inner]"
- **Many** statistical results are in lists
</div>

## Your Turn

Run this command: 

```{r}
example_list <- lm(mpg~wt, data = mtcars)
```

Now, find: 

- The structure of the list
- The element named 'coefficients'
- The fourth child element in the fifth parent element ('fitted.values')

## Solutions

```{r}
example_list <- lm(mpg~wt, data = mtcars)
str(example_list)
example_list$coefficients
example_list[[5]][4]
```


## Which() Function

- Usefull when subsetting
- Returns the index of any matching item

```{r}
which(letters[1:20] > "c")
```


## Your Turn

Using the mtcars data and the which function, 
find index of cars that have an MPG value greater than 25

Extra: What are the names of these cars? 

## Solution

```{r}
which(mtcars$mpg > 25)

rownames(mtcars[mtcars$mpg > 25,])
```


# 4. Functions

##

* Anytime you copy and paste more than twice, write a function
* Functions can take named and unnamed arguments
* Must be defined before called
* General structure outlined on next slide

##

```{r echo = T, eval = F}
newFunctionName <- function(manArg,optArg=1){
  if(optArg > 1) {
    optArg * 20
  } else { manArg * optArg  }
}
# Tests
newFunctionName(10) # = 10
newFunctionName(10,-1)# = -10
newFunctionName(10,3)# = 60
newFunctionName() # ERROR
```


## Your Turn

- Create a new function that: 
  * Takes 2 optional arguments (x=1,y=2)
  * adds the squar roots of the two arguments together
  
## Solution

```{r eval = F, echo= T}
newFunctionName <- function(x=1,y=2){
  sqrt(x)+sqrt(y)
}
#Tests
newFunctionName()
newFunctionName(3)
newFunctionName(5,6)
```

  